Domestic cooking and likelihood

### 2020-01-07 inst/cook.Rmd
changes in cooking strength

>>>>

## Changes in cooking strength.

We now consider a sequence of observations where one person, say
person 1, is suspected of an increase in strength (perhaps they
attended a domestic cookery course).  We make some observations before
and after the training and seek evidence that it was effective.
Unlike the previous cases, here we have a full order statistic without
the complication of not being allowed to vote for oneself (perhaps a
professional chef puts the dishes in order).  First, before:

```{r,beforetraining}
H <- hyper2(pnames=c(paste("p",1:5,sep=""),"t"))  # 't' the strength of training
# observations before training:
Hbefore <- H +
    order_likelihood(c(5,2,3,4,1)) + #  5 is the best, 4 the worst.
    order_likelihood(c(5,3,2,4,1)) + #  5 is the best, 1 the worst.
    order_likelihood(c(4,5,3,2,1)) + #  4 is the best, 1 the worst.
    order_likelihood(c(5,3,4,1,2)) + #  5 is the best, 2 the worst.
    order_likelihood(c(2,5,3,1,4))   #  2 is the best, 4 the worst.
Hbefore
```

Now, after the training we have two order observations: `1,2,3,4,5`
and `3,1,5,2,4`.  Thus player 1 comes first and second (after player
3) respectively.  The likelihood for the first would be

\[
\frac{p_1+t}{p_1+t+p_2+p_3+p_4+p_5}\cdot
\frac{p_2  }{      p_2+p_3+p_4+p_5}\cdot
\frac{p_3  }{          p_3+p_4+p_5}\cdot
\frac{p_4  }{              p_4+p_5}\cdot
\frac{p_5  }{                  p_5}\cdot
\]

and the likelihood for the second would be

\[
\frac{p_3  }{p_1+t+p_2+p_3+p_4+p_5}\cdot
\frac{p_1+t}{p_1+t+p_2+    p_4+p_5}\cdot
\frac{p_5  }{      p_2+    p_4+p_5}\cdot
\frac{p_2  }{      p_2+    p_4    }\cdot
\frac{p_4  }{              p_4    }\cdot
\]

We can translate these into package idiom by defining likelihood
functions `H1after` and `H2after` for the two observations:

```{r,aftertraining1}
H1after <- H
H1after[c("p1","t")] %<>% inc()
H1after[  "p2"     ] %<>% inc()
H1after[  "p3"     ] %<>% inc()
H1after[  "p4"     ] %<>% inc()
H1after[  "p5"     ] %<>% inc()
H1after[c("p1","t","p2","p3","p4","p5")] %<>% dec()
H1after[c(         "p2","p3","p4","p5")] %<>% dec()
H1after[c(              "p3","p4","p5")] %<>% dec()
H1after[c(                   "p4","p5")] %<>% dec()
H1after[c(                        "p5")] %<>% dec()
H1after
```

```{r,aftertraining2}
H2after <- H
H2after[  "p3"     ] %<>% inc()
H2after[c("p1","t")] %<>% inc()
H2after[  "p5"     ] %<>% inc()
H2after[  "p2"     ] %<>% inc()
H2after[  "p4"     ] %<>% inc()
H2after[c("p1","t","p2","p3","p4","p5")] %<>% dec()
H2after[c("p1","t","p2",     "p4","p5")] %<>% dec()
H2after[c(         "p2",     "p4","p5")] %<>% dec()
H2after[c(         "p2",     "p4"     )] %<>% dec()
H2after[c(                   "p4"     )] %<>% dec()
H2after
```

Combine:

```{r,combineboth}
(Htot <- Hbefore + H1after + H2after)  # observations are independent!
(ML_unconstrained <- maxp(Htot))
```

We see the training entity has an estimated strength of about 0.63,
higher than all real players combined.  At the evaluate the
loglikelihood is

```{r,loglikateval}
loglik(indep(ML_unconstrained),Htot)
```

We can test the hypothesis that the training was ineffective by
constraining the optimization to $t=0$:

```{r}
small <- 1e-4
(ML_constrained <- maxp(Htot,startp=rep(0.2*(1-small*2.1),5),fcm=c(1,1,1,1,1),fcv=1-2*small))
loglik(indep(ML_constrained),Htot)
```

So the support for our alternative hypothesis would be

```{r}
loglik(indep(ML_unconstrained),Htot)-loglik(indep(ML_constrained),Htot)
```

thus exceeding Edwards's two units of support.  We have strong
evidence to suggest that the training was effective.
